---
title: "`lumos`: Shed Light on Your Data"
author: "Benjamin Rich"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    css: vignette.css
    toc: true
vignette: >
  %\VignetteIndexEntry{`lumos`: Shed Light on Your Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{table1}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, comment='', render=knitr::normal_print)

library(lumos, quietly=TRUE)
```


The `lumos` package provides one main function, also called `lumos()` (but it can
be abbreviated simply to `l()` to avoid typing).  The main uses cases of this
function are to quickly explore data interactively in the console, or create
simple tabular summaries in R markdown documents.  Similar to `summary()`, but
aims to be as convenient as possible and produce nicer looking outputs.

If we call `lumos()` with a `data.frame` as its sole argument, it outputs a table
summarizing the variables, including the columns: `variable` (its name), `label`
(only present if at least one variable has a `label` atrribute), `class`,
`missing` (count) and `example` (a single value from that variable, typically
the first nonmissing value).  Let's try it on the Boston housing data from the
`MASS` package:

```{r}
library(survival)
l(pbc)


library(MASS)
lumos(Boston)

pima <- Pima.tr2
l(pima)


titanic <- as.data.frame(Titanic)
l(titanic)
l(titanic, Survived)
l(titanic, Class, Survived)
```

When called with `data` and one other argument, if the argument is
categorical outputs a frequency table and if it is continuous outputs a few
descriptive statistics (mean, standard deviation, median, min and max). The
`max` option is used to decide if a numeric argument is continuous or
categorical.

When called with more than one argument following `data`, those
arguments should all be categorical (`.max` is ignored in this case).
A frequency table is produced for the combinations of the categories, nested
from left to right. Percentages are not shown, just counts, and no sorting
is done (the categories appear in the order of factor levels).

By default, the function `knitr::kable` is used to format the
output so you get nice looking tables in both the console and in R markdown
documents.

If the `.gen` argument is `TRUE`, then something different
happens.  Instead of outputing a table, the function prints code statements
into the console: a call to `lumos()` for each variable in `data`.
The code can be copied from the console back into the script and used to
explore the `data.frame` one variable at a time. This is useful because
it saves the need to type the code for each variable.

